Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2020-06-29 ms/ms.tex
update ISF
a bit more update for Apr11 date

%%% Fig - tau summary %%%
\begin{figure}[b!]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Apr11.pdf}
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs is shown for different regions. The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers). In contrast, the effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although this is significant only for Germany (i.e., zero is not included in 75\% CI). 
	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date (\autoref{table:NPI_dates}). Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e. interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.) \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, rather then Apr~11, 2020.
    }
    \label{fig:tau-summary}
\end{figure}

@@

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant). The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}). 
When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
The exceptions are Austria, Belgium, Norway, and United Kingdom. % TODO UPDATE

@@

In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~2 ($\pm$2.85 days 95\% CI; \autoref{fig:late}). % TODO CHECK

@@

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).

Germany...

Sweden...

Denmark...

@@

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (i.e. free or fixed model in \autoref{table:WAIC}). %TODO check
It could be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as changing weather, news of new treatments, and international outbreaks.

@@

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Oren Kolodny for discussions and comments.
This work was supported in part by the Israel Science Foundation 552/19 and 1399/17.
}

>>>>

%%% Fig - tau summary %%%
\begin{figure}[b!]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Apr11.pdf}
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs is shown for different regions. The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers). In contrast, the effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although this is significant only for Germany (i.e., zero is not included in 75\% CI). 
	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date (\autoref{table:NPI_dates}). Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e. interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.) \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, rather then Apr~11, 2020.
    }
    \label{fig:tau-summary} %TODO maybe to put some footnote about the realibility of the result for Sweden and Denmark?
\end{figure}

@@

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant). The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).  
For all of the 11 European countries the out-of-sample root mean squared error (RMSE) is lower for the model with a free~$\tau$ (\autoref{table:RMSE}). We don't check it for Wuhan, China, because we have no out-of-sample data for it.  % TODO reference? ; add table ; find more data for Wuhan?
When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in all of 12 countries except Sweden and Denmark. Both Sweden and Denmark, don't have one visual peak in daily cases for the dates the models are evaluated, so probably it is not enough data for the free~$\tau$ model.

@@

In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~3 ($\pm$2.85 days 95\% CI; \autoref{fig:late}). % TODO CHECK

@@

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
Germany... %TODO
Sweden and Denmark show early effective start of NPIs. However, for both of the contries, the dataset that we use doesn't contain dates with the peak of the pandemic. WAIC model comparison method doesn't prefer the model with a free~$\tau$ on the model with a fixed~$\tau$.
Therefore we can't trust in the outcome of the free~$\tau$ model for those countries.

@@

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 10 out of 12 regions (i.e. free or fixed model in \autoref{table:WAIC}). 
It could be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as changing weather, news of new treatments, and international outbreaks.

@@

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Oren Kolodny for discussions and comments.
This work was supported in part by the Israel Science Foundation 3811/19 and 1399/17.
}
